---
output: html_document
editor_options: 
  chunk_output_type: console
---
White wine data EDA 
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}
library(ggplot2)
library(GGally)
library(scales)
library(memisc)
library(gridExtra)
library(grid)
library(dplyr)
library(psych)
```



```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd('C:/Users/Xuefei Yu/Desktop/Udacity/EDA_R')
wines <- read.csv('wineQualityWhites.csv')
head(wines)

# remove the first column of wines dataset

wines$X <- NULL

dim(wines)
names(wines)
str(wines)
summary(wines)
```

This wine dataset consists of 13 features, with 4898 observations, and was 
created by: Paulo Cortez (Univ. Minho), Antonio Cerdeira, Fernando Almeida, 
Telmo Matos and Jose Reis in 2009. This dataset is a tidy dataset, and 
contains no missing values. The 13 features are: X, fixed.acidity,volatile.
acidity, citric.acid, residual.sugar, chlorides, free.sulfur.dioxide, 
total.sulfur.dioxide, density, pH, sulphates, alcohol and quality.The first 
column or feature X is index that can be removed from the original dataset.

# Univariate Plots Section



```{r}
#since I have to plot several histograms for each feature, I create this 
#function to make things easier. Function arguments: a <- data feature, 
# c <- binsize, s <- historam x axis break sequence, l <- histogram x axis
# limits
histogramplot <- function(a,c = NULL , s  ,l ){
if (missing(s) & missing(l) ) {
 ggplot(aes(x = a), data = wines) +
  geom_histogram(binwidth = c, colour="black", fill="blue") +
  scale_x_continuous() +
   xlab(deparse(substitute(a))) 
}
else if (missing(l)) {
ggplot(aes(x = a), data = wines) +
  geom_histogram(binwidth = c, colour="black", fill="blue") +
  scale_x_continuous(breaks = s) +
   xlab(deparse(substitute(a)))  
}

else if (missing(s)) {
ggplot(aes(x = a), data = wines) +
  geom_histogram(binwidth = c, colour="black", fill="blue") +
  scale_x_continuous(limits = l) +
   xlab(deparse(substitute(a)))  
}  
else  
  
ggplot(aes(x = a), data = wines) +
  geom_histogram(binwidth = c, colour="black", fill="blue") +
  scale_x_continuous(breaks = s, limits = l) +
   xlab(deparse(substitute(a)))
 
} 

```


```{r}

histogramplot(wines$quality, 1, s = seq(3, 9, 1))

summary(wines$quality)
```

Quality distribution is generally a normal distribution.

```{r}

histogramplot(wines$fixed.acidity)


histogramplot(wines$fixed.acidity,0.1)

summary(wines$fixed.acidity)

```

Most wines have the fixed acidity between 5 to 9 g / dm^3.

```{r}

histogramplot(wines$volatile.acidity)

histogramplot(wines$volatile.acidity,0.01)

summary(wines$volatile.acidity)
```


Most wines have the volatile acidity between 0.15 to 0.45 g / dm^3.

```{r echo=FALSE}

histogramplot(wines$citric.acid)

histogramplot(wines$citric.acid, 0.01,seq(0, 1.7, 0.1), c(0,0.75))

summary(wines$citric.acid)
```

The citric acid follow a normal distribution with most values between 
0.1 to 0.6 g / dm^3. Then changing the bin size to 0.01 and limit x value to 
0.75, so we can see more details about this variable. A great amount of wines 
have citric value at 0.49, this maybe caused by rounded values.

```{r echo=FALSE}
wines$total.acidity <- wines$fixed.acidity + wines$volatile.acidity + 
                      wines$citric.acid

histogramplot(wines$total.acidity)

histogramplot(wines$total.acidity, 0.1, seq(4, 11, 1), c(4,11))

summary(wines$total.acidity)
```

By reading online source, it says that the total acidity is the amount of fixed 
acidity plus the volatile acidity. So I create a new variable by sum of all 
the fixed, volatile and citricacids variables.

```{r echo=FALSE}

histogramplot(wines$residual.sugar, 1)

ggplot(aes(x = residual.sugar), data = wines) +
  geom_histogram(binwidth = 0.02, colour="black", fill="blue") +
  scale_x_log10(breaks = c(1, 2, 5, 10, 20))
summary(wines$residual.sugar)
```

The residual sugar distribution is skewed distribution. Transformed the long 
tail data to better understand the distribution of residual sugar. T
he tranformed residual sugar distribution appears bimodal with the price peaking
around 1.4 or so and again at 9 or so. 

```{r echo=FALSE}

histogramplot(wines$chlorides)

histogramplot(wines$chlorides,0.005,l= c(0,0.1))

summary(wines$chlorides)
```

Limited variable value to 0.1 to eliminate some outliers. And Chlorides follows
normal distribution. Most chlorides are between 0.025 and 0.075.

```{r echo=FALSE}
histogramplot(wines$sulphates)

histogramplot(wines$sulphates,0.01,seq(0.2, 1.1, 0.1))

summary(wines$sulphates)

```

Most sulphates values are between 0.3 to 0.7 g / dm^3.

```{r echo=FALSE}

histogramplot(wines$alcohol)

histogramplot(wines$alcohol, 0.1,seq(8, 15, 0.5))

summary(wines$alcohol)
```

Most alcohol values are 9.5 to 13 % by volume. The distribution is right skewed 
distribution.

```{r echo=FALSE}

histogramplot(wines$pH)

summary(wines$pH)

```

pH follow a normal distribution with mean around 3.2

```{r echo=FALSE}

histogramplot(wines$density)

histogramplot(wines$density,0.0005,l = c(0.985,1.005))

summary(wines$density)

```

Limited variable value from 0.985 to 1.005 to eliminate some outliers. 
And density follows normal distribution. Most densities are between
0.990 and 1.000.

```{r echo=FALSE}

histogramplot(wines$total.sulfur.dioxide)

histogramplot(wines$total.sulfur.dioxide,5,seq(0,275, 25),c(0,275))

summary(wines$total.sulfur.dioxide)

```

Limited variable value from 0 to 275 to eliminate some outliers. 
And it follows normal distribution. Most variables are between
50 and 250.

```{r echo=FALSE}

histogramplot(wines$free.sulfur.dioxide)

histogramplot(wines$free.sulfur.dioxide,1,seq(0, 290, 10),c(0,90))

summary(wines$free.sulfur.dioxide)

```

The free sulfur dioxide follow a normal distribution with mean near 35.

# Univariate Analysis

### What is the structure of your dataset?

There are 4898 observations of wine in this dataset with 12 features 
(fixed acidity, volatile acidity, citric acid, residual sugar, chlorides, 
free sulfur dioxide, total sulfur dioxide, density, pH, sulphates, alcohol 
and quality). Quality is the output attribute that we are interested in.
All the features are numerical.

### What is/are the main feature(s) of interest in your dataset?

The main feature that I want to learn and predict is the quality.
I would like to determine which features are best for predicting the quality of 
wine. I suspect density, alcohol,residual sugar and some combination of 
acidities can be used to build a predictive model to price diamonds.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Total sulfur dioxide and density likely contribute to wine quality.

### Did you create any new variables from existing variables in the dataset?

Sum of all acidities is what I create so far. I may create some combinations 
later.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

1. Quality has most values concentrated in the categories 5, 6 and 7.
Only a small proportion is in the categories [3, 4] and [8, 9] and none in the 
categories [1, 2] and 10.

2. There are outliers in Fixed acidity, volatile acidity and citric acid. 
I eliminated these outliers, and distribution of these variables seems to be 
more symmetric.

3. Residual sugar has a right skewed distribution, and it remain right skewed 
after taking out outliers.

4. Alcohol has an anormal distribution, but it does not have outliers.


# Bivariate Plots Section
```{r echo=FALSE, Univariate_Plots}

ggcorr(wines, label = T, layout.exp = 2,nudge_x = -1)
```

Pairwise plotting variables using 'ggcorr' function to get a general 
correlation sense of data.

```{r echo=FALSE, Bivariate_Plots}
wines$quality.factor <- factor(wines$quality)
```

```{r echo=FALSE}

tmpdata1 <- subset(wines, 
             select = -c(fixed.acidity, volatile.acidity, citric.acid, quality))

pairs.panels(tmpdata1,pch = '.')

```

Faceted all paired features to genarally learn relation of them. 

```{r echo=FALSE}

tmpdata2 <- subset(tmpdata1, 
                select = -c(free.sulfur.dioxide, pH,sulphates))
pairs.panels(tmpdata2, pch=".")
```

For this data, the most interesting part is what determine the wine quality. 
So I ingnore some features which are not quite related with quality variable.
The features I'm interested in are: residual.sugar, chlorides, 
total.sulfur.dioxide, density, alcohol and total.acidity.

```{r echo=FALSE}

ggplot(aes (x = quality.factor, y = alcohol ), data = wines) +
  geom_boxplot()

```

Then I factor quality feature to boxplots with these related features. From this 
alcohol vs quality plot, wines with quality greater than 6 generally have higher 
alcohol content.

```{r echo=FALSE}
ggplot(aes (x = quality.factor, y = density ), 
       data = subset(wines, wines$density < 1.005))  +
  geom_boxplot()

```

From this density vs quality plot, wines with quality greater than 6 generally 
have lower density. Since alcohol density is less than water, wines with more
alcohol have lower density and higher quality.

```{r echo=FALSE}
ggplot(aes (x = quality.factor, y = residual.sugar ), 
       data = subset(wines, wines$residual.sugar < 40)) +
  geom_boxplot()

```

From this residual suagr vs quality plot, it is hard to tell any relation in it.
But the highest quality wine have the slightest residual residual sugar.

```{r echo=FALSE}
ggplot(aes (x = quality.factor, y = chlorides ), 
       data = subset(wines, wines$chlorides < 0.1)) +
  geom_boxplot()

```

From this chlorides vs quality plot, it just have a little downward trend when 
quality inceasing.

```{r echo=FALSE}
ggplot(aes (x = quality.factor, y = total.acidity ), data = wines) +
  geom_boxplot()

```

From this total acidity vs quality plot, I couldn't see any relation from this 
boxplot.

```{r echo=FALSE}
ggplot(aes (x = quality.factor, y = total.sulfur.dioxide ), data = wines) +
  geom_boxplot()
```

I couldn't any relation from this boxplot. All the above boxplots are some how 
related to quality. Next, I will explore some other interesting features.   

```{r echo=FALSE}
ggplot(aes(x = residual.sugar, y = density), data = wines) +
  geom_point(color = I('orange'),alpha = 0.5, size = 1, position = 'jitter') +
  coord_cartesian(xlim = c(0, quantile(wines$residual.sugar, .95))) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle('Density by Residual Sugar')
```

Residual sugar and density have 0.84 correlation. This is verified from this 
plot. I think this is because sugar solvent has higher density.

```{r echo=FALSE}
ggplot(aes(x = density, y = alcohol), data = wines) +
  geom_point(color = I('orange'), alpha = 0.5, size = 1, position = 'jitter') +
  coord_cartesian(ylim = c(quantile(wines$alcohol, .01),
                           quantile(wines$alcohol, .99))) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle('Alcohol by Density')
```

Also,  a strong relationship between density and alcohol is verified. Their 
correlation is -0.78.

```{r echo=FALSE}

ggplot(aes(x = alcohol, y = quality), data = wines) +
  geom_point(color = I('orange'), alpha = 0.5, size = 2, position = 'jitter') +
  coord_cartesian(xlim = c(quantile(wines$alcohol, .01),
                           quantile(wines$alcohol, .99))) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle('Quality by Alcohol')

```

From alcohol and quality boxplot, I thought they should have a relation. So I 
made this scatter plot to verify it. 

```{r echo=FALSE}
ggplot(aes(x = density, y = quality), data = wines) +
  geom_point(color = I('orange'), alpha = 0.5, size = 2, position = 'jitter') +
  coord_cartesian(xlim = c(quantile(wines$density, .01),
                           quantile(wines$density, .99))) +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle('Quality by Density')


```

Also, I made this quanlity VS density scatter plot to verify my thought from 
their boxplot.


# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I observed quality vs alcohol and quality vs density since quality is the main
interest of this dataset. Alcohol has 0.44 correlation wieh quality while 
density has -0.31. But alcohol and density are somehow related because alcohol 
has smaller densidy than water.

For other features, I couldn't any significant relations with quality.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Density generall has relation with all ingredients(acid, sugar, sodium). 

### What was the strongest relationship you found?

The strongest relationship is between density and residual sugar(0.84). 
The second one is density with alcohol (-0.78) .

# Multivariate Plots Section


```{r echo=FALSE, Multivariate_Plots}

ggplot(aes(x = density, y = alcohol, color = quality.factor), data = wines) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  coord_cartesian(xlim = c(quantile(wines$density, .01),
                           quantile(wines$density, .99))) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~quality.factor) +
  scale_color_brewer(type = 'div', 
                     guide = guide_legend(title = 'Color', reverse = T,
                                  override.aes = list(alpha = 1, size = 2))) +
                                  theme_dark() +
  ggtitle('Alcohol vs Density by Quality')

```

Facet Alcohol vs Density scatter plots by Quality. I can see the distribution 
move from bottom-right to top-left when quality increasing.  

```{r echo=FALSE}


ggplot(aes(x = density, y = alcohol , colour = quality.factor), data = wines) +
  geom_point(alpha = 0.8, position = 'jitter') +
  coord_cartesian(xlim = c(quantile(wines$density, .01), 
                           quantile(wines$density, .95)),
                  ylim = c(quantile(wines$alcohol, .01), 
                           quantile(wines$alcohol, .95))) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = T,
                                  override.aes = list(alpha = 1, size = 2))) +
                                  theme_dark() +
  ggtitle('Alcohol vs Density by Quality')


```

Scatter plot Alcohol vs Density scatter plots by Quality. I can clearly see the
negative correlation between Alcohol and Density, and the  positive 
correlation between Quality and Alcohol.

```{r}
wines$alcohol_bucket <- cut(wines$alcohol,
                             c(7.9, 9.0, 10.0, 11.0, 12.0, 13.0 ,14.0, 15.0))

table(wines$alcohol_bucket, useNA = 'ifany')

ggplot(aes(x = quality, y = density),
       data =subset(wines, wines$density < 1.005) ) +
  geom_line(aes(color = alcohol_bucket), 
           size = 2, stat = 'summary', fun.y = mean)+ 
  ggtitle('Density vs Quanlity by Alcohol bucket')


```

Made alcohol feature bucket c(7.9, 9.0, 10.0, 11.0, 12.0, 13.0 ,14.0, 15.0), and 
used it to plot mean of Density VS Quality. I can clearly see the relation 
between density and alcohol as I talked above. 

```{r echo=FALSE}

ggplot(aes(x = density, y = residual.sugar, colour = quality.factor),
       data = wines) +
  geom_point(alpha = 0.5, size = 1, position = 'jitter') +
  coord_cartesian(xlim = c(quantile(wines$density, .01),
                           quantile(wines$density, .99))) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~quality.factor) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = T,
                                  override.aes = list(alpha = 1, size = 2))) +
                                  theme_dark()+
  ggtitle('Residual Sugar vs Density by Quality')


```

Facet Residual Sugar vs Density scatter plots by Quality. I can see the
distribution move from bottom-right to bottom- left when quality increasing.

```{r echo=FALSE}
ggplot(aes(x = residual.sugar, y = alcohol, colour = quality.factor), 
       data = wines) +
  geom_point(alpha = 0.8, position = 'jitter') +
  coord_cartesian(xlim = c(0, quantile(wines$residual.sugar, .95))) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = T,
                                  override.aes = list(alpha = 1, size = 2))) +
                                  theme_dark() +
  ggtitle('Alcohol vs Residual Sugar by Quality')

```

Scatter plot Alcohol vs Residual sugar by Quality. I can see the quality 
distribution when alcohol increasing. 

```{r echo=FALSE}
ggplot(aes(x = total.acidity, y = pH , colour = quality.factor), data = wines) +
  geom_point(alpha = 0.8, position = 'jitter') +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Color', reverse = T,
                                  override.aes = list(alpha = 1, size = 2))) +
                                  theme_dark()
  ggtitle('PH vs total.acidity by Quality')
```

Since acid should be related with PH value, I want to investgate if 
PH related to quality. From this plot, I can see the negative relation between
PH and Total.acidity. But I couldn't see any relation between PH and Quality.

```{r echo=FALSE}
m1 <- lm(quality ~ residual.sugar, data = wines)
m2 <- update(m1, ~ . + chlorides )
m3 <- update(m2, ~ . + alcohol)
m4 <- update(m3, ~ . + total.sulfur.dioxide)
m5 <- update(m4, ~ . + density)
m6 <- update(m5, ~ . + total.acidity)
mtable(m1, m2, m3, m4 ,m5, m6)
summary(m6)

```

Fit linear model using residual sugar, chlorides, alcohol, total.sulfur.dioxide,
density and total.acidity predictor.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In this multivariate analysis section, I kept further investigating relation in
alcohol, residual sugar, density and output feature quality. All these plots 
strenthened my thought from section 2.


### Were there any interesting or surprising interactions between features?

Acids and PH value are related, but both of them are not related to Quality.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I built a linear model trying to predict by useing 6 features: residual sugar, 
chlorides, alcohol, total.sulfur.dioxide,density and total.acidity. But I only 
got 'Multiple R-squared:  0.2145,	Adjusted R-squared:  0.2135' which is not
accurate enough.
------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}

ggplot(aes(alcohol), data = wines) +
  geom_histogram(binwidth = 0.1, colour="black", fill="blue") +
  scale_x_continuous(breaks = seq(8, 15, 0.5)) +
  xlab('alcohol (% by volume)')+
  ylab('wine counts') +
  ggtitle('Alcohol Feature Histogram')

summary(wines$alcohol)


```

### Description One

The distribution of alcohol is a little bit right skewed. This feature has no 
outliers. Most of wines have less than 11.0 (% by volume) alcohol. 

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes (x = quality.factor, y = alcohol), data = wines) +
  geom_boxplot(colour = 'black', fill = 'lightblue') +


  xlab('Wine Quality')+
  ylab('Alcohol (% by volume)') +
  ggtitle('Alcohol vs Quality Boxplot')
```

### Description Two

For wines' quality greater than 6, a increasing trend can be observed. But it is 

difficult to tell any relation for wines' quality less than 6.

### Plot Three
```{r echo=FALSE, Plot_Three}
ggplot(aes(x = density, y = alcohol , colour = quality.factor), data = wines) +
  geom_point(alpha = 0.8, position = 'jitter') +
  coord_cartesian(xlim = c(quantile(wines$density, .01), 
                           quantile(wines$density, .95)),
                  ylim = c(quantile(wines$alcohol, .01), 
                           quantile(wines$alcohol, .95))) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_brewer(type = 'div',
                     guide = guide_legend(title = 'Quality', reverse = T,
                                  override.aes = list(alpha = 1, size = 2))) +
                                  theme_dark()+
                                  
  ggtitle('Alcohol vs Density by Quality')+
  xlab('Wines Density (g / cm^3)')+
  ylab('Alcohol(% by volume)') 

```

### Description Three

The negative correlation between alcohol and density is clearly observed. Since 
the mid quality is in color white, it is easier to observe that higher quality 
wines tend to have more alcohol, and vise versa.

------

# Reflection

There are 4898 wines infomation in this wine dataset. Before I investigate this 
dataset, I managed to acquire some knowledge from internet to deeply understand 
these features in it.Basicly, Sweet Taste (sugars + alcohols) and Acid 
Taste(acids) determine the quality of wines.Then I create histogram for each 
feature to understand features. The main interest of this dataset is quality. 
I found several features might have relations with it: alcohol, sugar, density,
acid,etc. The most interesting feature for me is alcohol. Then I focused on 
relations in main interest,alcohol and other features. 

I found density almost has strong relation to all features. This is reasonable 
since all content in wines can affect density. I also found PH value is highly
related with acid. The most interesting part I found is high quality wines tend 
to have high content of alcohol.

Then I built a linear model to predict. Unfortunately, the accuracy is not good
enough. I will try to build some other regression model(e.g., support vector 
machine(SVM)) to enhance the accuracy. 




# Bibliography

https://en.wikipedia.org/wiki/Sodium_chloride#Chemistry

http://www.calwineries.com/learn/wine-chemistry